---
title: "Group Lab 2"
author: "Aldrin Carbonell, Johnny Nguyen, Justin Choi, Samuel Perng, Andy Tseng (Group 1)"
date: "February 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
# Loading data sources
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(raster)
library(rgdal)
library(plyr)
library(sf)
library(dplyr)
library(rnaturalearth) 
library(geosphere)
library(leaflet)
library(RColorBrewer)
library(networkD3)
library(tmap)
library(spdep)
library(ggplot2)
library(maptools)
library(ggmap)
library(spData)
detach("package:plyr", unload=TRUE)

# WAC, RAC, and OD data
odData = read.csv("data/wa_od_main_JT00_2015.csv.gz") %>% as_tibble()
wacData = read.csv("data/wac2015.csv") %>% as_tibble()
racData = read.csv("data/rac2015.csv") %>% as_tibble()
KCtract = read.csv("data/KC_Tracts.csv") %>% as_tibble()
#shapefile KC counties
KC_shape <- shapefile("data/shapefiles/king10ct.shp")
#Shapefile for KC blocks
KC_block <- shapefile("data/shapefiles/kc_block_10.shp")


```

```{r, include=FALSE}
# Part 2: Data Wrangling boogaloo

#TRACT LEVEL DATA---------------------------------------------------------------
# Converting GeoID columns to characters 
odData[, 1:2] <- sapply(odData[, 1:2], as.character)
wacData[1] <- sapply(wacData[,1], as.character)
racData[1] <- sapply(racData[,1], as.character)

#Selecting out tracts of interest non spatially (King County Tracts) ------------
# Subsetting the census tract (that we'll use to merge rows togethers)
odData$wTract <- substr(odData$w_geocode, 6, 11)
odData$hTract <- substr(odData$h_geocode, 6, 11)
#Work area subset
wacData$wTract <- substr(wacData$w_geocode, 6, 11)
#Residential area Subset
racData$hTract <- substr(racData$h_geocode, 6, 11)

#OD Aggregated by Tracts
ODAGtract=aggregate(. ~ odData$wTract + odData$hTract,data=odData[,c(3:12)],FUN=sum)
#Wac Aggregated by Tracts
WACAGtract=aggregate(. ~ wacData$wTract,data=wacData[,c(2:52)],FUN=sum)
#Rac Aggregated by Tracts
RACAGtract=aggregate(. ~ racData$hTract,data=racData[,c(2:42)],FUN=sum)

#finding tracts that are within King County from a file that only has KC tracts
KCtracts <- dplyr::select(KCtract, GEOID10)
KCtracts$GEOID10 <- substr(KCtracts$GEOID10, 6, 11)


#Selecting tracts where the work and home tracts are within King County for the OD data
KCodtract <- filter(ODAGtract, ODAGtract$`odData$wTract` %in% KCtracts$GEOID10 == TRUE,
                      ODAGtract$`odData$hTract` %in% KCtracts$GEOID10 == TRUE)
KCodtract <- mutate(KCodtract, GEOID10 = paste0('53033', KCodtract$`odData$hTract`))

#Flatten COlumn to 1D Vector
WACAGtract$`wacData$wTract` <- WACAGtract$`wacData$wTract`[,1]
RACAGtract$`racData$hTract` <- RACAGtract$`racData$hTract`[,1]


#Selecting tracts in King County for aggregated WAC data
KCwactract <- filter(WACAGtract, WACAGtract$`wacData$wTract` %in% KCtracts$GEOID10 == TRUE) 
KCwactract <-  mutate(KCwactract, GEOID10 = paste0('53033', KCwactract$`wacData$wTract`))
#Selecting tracts in King County for aggregated RAC data
KCractract <- filter(RACAGtract, RACAGtract$`racData$hTract` %in% KCtracts$GEOID10 == TRUE)
KCractract <-  mutate(KCractract, GEOID10 = paste0('53033', KCractract$`racData$hTract`))

#Joining Data to tract level shapefiles -----------------------------------------

#Joining rac with the shapefile
KC_rac_with_shape <- merge(KC_shape, KCractract, by='GEOID10')
#Joining wac with the shapefile
KC_wac_with_shape <- merge(KC_shape, KCwactract, by='GEOID10')



#BLOCK LEVEL DATA---------------------------------------------------------------

#OD data by block in KC
ODblk <- filter(odData, odData$w_geocode %in% KC_block$GEOID10 == TRUE,
                odData$h_geocode %in% KC_block$GEOID10 == TRUE)
#Work area characteristics by block in KC
WACblk <- filter(wacData, wacData$w_geocode %in% KC_block$GEOID10 == TRUE)
WACblk <- mutate(WACblk, MERGEID = WACblk$w_geocode)
#Residential area characteristics by block in KC
RACblk <- filter(racData, racData$h_geocode %in% KC_block$GEOID10 == TRUE)
RACblk <- mutate(RACblk, MERGEID = RACblk$h_geocode)
#Join to block level shapefiles
WACblk_with_shape <- merge(KC_block, WACblk, by.x='GEOID10', by.y='MERGEID')
RACblk_with_shape <- merge(KC_block, RACblk, by.x='GEOID10', by.y='MERGEID')
```


# Mapping Origin-Destination Data on Employment and Industry in King County, Washington 
<!-- Step 1: intro/context -->

This data was gathered under the US Census Bureau. These datasets are known as "LODES" , which stands for LEHD (Longitudinal Employer-Household Dynamics) Origin-Destination Employment Statistics. That's quite an abbreviation, but it's basically information about occupational and residential data.
```{r include=FALSE}
#Step 2: non-spatial visualizations
#visualize OD RAC WAC Tracts 
#-----------------------OD BAR-----------------------------
#Select top 10 Tracts with the highest Total Number of Jobs
topJobsOD <- KCodtract %>% top_n(10, S000)
#Change Column Names
names(topJobsOD)[1] <- "wTract"
names(topJobsOD)[2] <- "hTract"
#Add New Column Representing the Home to Work Block Code
topJobsOD <- topJobsOD %>% mutate(
                 homeToWork = paste(hTract,wTract, sep = '-'))
#Plot 
odBar <- ggplot(data=topJobsOD, aes(x=homeToWork, y=S000)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y="Total Number of Jobs", x = "Home Census Tract Code to Work Census Tract Code") +
  ggtitle("Top 10 Census Tracts by Total Number of jobs - OD") 

#----------------------------WAC BAR--------------------------
#Select top 10 Tracts with the highest Total Number of Jobs
topJobsWAC<- KCwactract %>% top_n(10, C000)
#Chgange Column Names
names(topJobsWAC)[1] <- "wTract"
#Plot
wacBar<-ggplot(data=topJobsWAC, aes(x=wTract, y=C000)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y="Total Number of Jobs", x = "Census Tract Code") + 
  ggtitle("Top 10 Census Tracts by Total Number of jobs - WAC") 

#-------------------------------RAC---------------------------
#Select 10 10 Records with Total Number of Jobs
topJobsRAC<- KCractract %>% top_n(10, C000)
#Change Column Names
names(topJobsRAC)[1] <- "hTract"
#Plot
racBar<-ggplot(data=topJobsRAC, aes(x=hTract, y=C000)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y="Total Number of Jobs", x = "Census Tract Code") +
  ggtitle("Top 10 Census Tracts by Total Number of jobs - RAC")

##-----------------------MAPPING OD----------------------------
#Store ODtract records into new tibble
KCodtract2 <- KCodtract
#Change Column names
names(KCodtract2)[1] <- "wTract"
names(KCodtract2)[2] <- "hTract"
#Subsetting data by the tract that has the most jobs
KC_od_with_shape1 <- merge(KC_shape, subset(KCodtract2, wTract == "000200"), by='GEOID10')
KC_od_with_shape2 <- merge(KC_shape, subset(KCodtract2, wTract == "010100"), by='GEOID10')
KC_od_with_shape1 <- st_as_sf(KC_od_with_shape1)
KC_od_with_shape2 <- st_as_sf(KC_od_with_shape2)
#mapping  test
map2 <- ggplot(data=KC_od_with_shape1) +
  geom_sf()+
  geom_sf(data=subset(KC_od_with_shape1, wTract == "000200"), aes(fill="red")) + 
  geom_sf(data=subset(KC_od_with_shape1, hTract == "000200"), aes(fill="blue")) + 
  scale_fill_discrete("Census Tract", 
                      labels=c("Destination", "Home"))

#-------------------------- Home to work Visualization 1 ----------------------------------
##visualizing home to work
#Subset data to be to the top tracts
flowdata <- subset(KCodtract2, wTract == "000200") %>%
  subset(hTract != "000200") %>% dplyr::select(hTract, wTract,S000)
#Choose top 10 records
flowdata <- flowdata %>% top_n(10, S000)
#Store tracts into vector
censusTracts <- c(flowdata$hTract, unique(flowdata$wTract))
nodes <- data.frame(name = censusTracts, id = 0:(length(censusTracts)-1))
links <- flowdata %>%
  left_join(nodes, by = c('hTract' = 'name')) %>%
  rename(origin_id = id) %>%
  left_join(nodes, by = c('wTract' = 'name')) %>%
  rename(dest_id = id)

#Plot Visualization
HWV <-sankeyNetwork(Links = links, Nodes = nodes, Source = 'origin_id', Target = 'dest_id', 
              Value = 'S000', NodeID = 'name', fontSize = 16)

#----------VISUALISING HOME TO WORK ON A MAP----------------------------------------------
#Subset data to get to 000200 the Census tract with the most total number of jobs
mappingdata <- merge(KC_shape, subset(KCodtract2, wTract == "000200"), by='GEOID10')
mappingdata <- st_as_sf(subset(mappingdata, wTract=="000200"))

#Getting coordinates for Census Tract 2000
mappingdata <- mappingdata %>% dplyr::select(wTract, hTract, 11,12,S000)
mappingdata2 <- as.tibble(mappingdata %>% subset(hTract=="000200") %>% dplyr::select(3,4))
mappingdata$work.lat <- as.character(mappingdata2[1,1]) 
mappingdata$work.long <- as.character(mappingdata2[1,2])  

#Changing column types to work with functions
MP <- as.tibble(mappingdata %>% top_n(10, S000))
MP$work.lat <- as.numeric(as.character(MP$work.lat))
MP$work.long <- as.numeric(as.character(MP$work.long))
MP$INTPTLAT10 <- as.numeric(as.character(MP$INTPTLAT10))
MP$INTPTLON10 <- as.numeric(as.character(MP$INTPTLON10))

#Get Lines for Plot that go from Home to Work
flows <- gcIntermediate(MP[,4:3], MP[,7:6], sp = TRUE, addStartEnd = TRUE)
flows$counts <- MP$S000
flows$origins <- MP$hTract
flows$destinations <- MP$wTract

#Hover Overlay
hover <- paste0(flows$origins, " to ", 
                flows$destinations, ': ', 
                as.character(flows$counts))
pal <- colorFactor(brewer.pal(4, 'Set1'), flows$origins)

#Save to Map
odMAP<-leaflet() %>%
  addProviderTiles(providers$Stamen.Toner) %>%
  addPolylines(data = flows, weight = (flows$counts)*.02, label = hover, 
               group = ~origins, color = ~pal(origins)) %>%
  addLayersControl(overlayGroups = unique(flows$origins), 
                   options = layersControlOptions(collapsed = FALSE))


```
## OD, WAC, and RAC Bar Graphs
```{r echo=FALSE}
#Step 2
odBar
```
```{r echo=FALSE}
wacBar
```
```{r echo=FALSE}
racBar
```

  Looking at the first graph that represents the top Origin to Destination relationships, We saw that 000200 was a popular place to work. One thing that stuck out in the bar graph was a lot of people who live in 000200 also work in 000200. So we wanted to see what other home census tracts also travel to 000200. 

## Aggregated Block and Tract Level Data for King County
For the two static maps below, we chose to visualize the amount of jobs in each block and tract that require employees to have at least a Bachelor's degree. In the LODES dataset, this variable is denoted as **CD04**.
```{r echo=FALSE}
# Step 3: Mapping------------------------------------------------------------------
# map for block lvl
tm_shape(RACblk_with_shape) + 
  tm_fill(title = "Number of jobs for workers  \n that have attained at least a \n Bachelor's degree \n (by Census Blocks)",
          col = "CD04") +
  tm_legend(legend.outside = TRUE) +
  tm_borders(lwd = 1, alpha = 0.04)
```

As you can see, the map above that represents block level data is far more disorganized than the the map below, which represents tract level data. Since King County is a large scale area relative to a single census block, the variable in this map is difficult to visualize at a full extent. Census tracts are larger than census blocks, making it easier to view patterns.
```{r echo=FALSE}
# map for tract lvl
tm_shape(KC_rac_with_shape) + 
  tm_fill(title = "Number of jobs for workers  \n that have attained at least a \n Bachelor's degree \n (by Census Tracts)",
          col = "CD04", n = 6) +
  tm_legend(legend.outside = TRUE) +
  tm_borders()
```
  
## Origin-Destination Flow Map
  Below is a flow visualization that represents the Origin to Destination relationship. These census tracts are the top 10 census tracts with the most number of jobs to 000200. Hovering over the lines will give you the number of jobs the Origin to Destination relationship has.  Let's see what this looks like on an actual map. 
```{r echo=FALSE}
# Step 4
HWV
```
## Origin-Destination Interactive Leaflet Map
Here is a map that represents the above visualization with respect to it's location. It looks like most people who live in downtown Seattle also work in this area. 
```{r echo=FALSE}
#Step 5
odMAP
```

```{r echo=FALSE}
#Subset data so that there are no NA values in a column
KC_wac_with_shape2 <- subset(KC_wac_with_shape,!is.na(CE01))
#Create map for Column CEO1 - Number of jobs earning less than $1250/month
lowEarningsMap <-tm_shape(KC_wac_with_shape2) + tm_polygons(style="quantile", col = "CE01") +
  tm_legend(outside = TRUE, text.size = .8) 
#Getting Weights for Morans Statistics
nb <- poly2nb(KC_wac_with_shape2, queen=TRUE)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
KC_wac_with_shape2$CE01.lag <- lag.listw(lw, KC_wac_with_shape2$CE01)
MC<- moran.mc(KC_wac_with_shape2$CE01, lw, nsim=599)
# calculate the local moran 
lmoran <- localmoran(KC_wac_with_shape2$CE01, lw)
```

## Calculating Moran's Statistic
CE01 Represents the Number of jobs with earnings $1250/month or less 
As we see hee there seems to be a spatial correlation between Number of jobs
earnings less than $1250/month. To see if there is a Spatial Correlation we calculated 
Moran's Statistic. Since the Moran's Statistic is positive (indicated by the vertical bar) we
can assume that there is a spatial correlation for Number of jobs Earning Less than $1250/month
```{r echo=FALSE}
#Step 6
lowEarningsMap
plot(MC, main="", las=1)
summary(lmoran)
```



